HMMotion: Using tracking data to predict coverage strategies in American Football

### 2025-01-03 submission.Rmd
updating HMM parts

#### 2. Hidden Markov model
We describe the movements of the five defensive players during the phase of pre-snap motion within a hidden Markov model (see the Appendix for a thorough description). In particular, we assume the defender's $y$-coordinates in each time step $t$ to be realizations of Gaussian distributions with means according to the $y$-coordinates of the offensive players to be guarded (these are the underlying states) and a standard deviation to be estimated (see Franks et al. 2015 for a similar approach in basketball).

This probabilistic approach allows us to infer dynamic coverage patterns during plays with pre-snap motion.

OLE

@@

@@

#### Hidden Markov Model
A hidden Markov models consists of an observed time series $\{\boldsymbol{y}_t\}_{t=1}^T$ --- here, the y-coordinates of the defensive players and an unobserved first-order Markov chain $\{ g_t\}_{t=1}^T$, with $g_t \in \{1,\ldots,N\}$ which proxies the offensives players to be guarded at every time point $t$. The Markov chain is fully described by an initial distribution $\boldsymbol{\delta}=\bigl( \Pr(g_1=1), \ldots, \Pr(g_1=N) \bigr)$ and a transition probability matrix (t.p.m.) $\boldsymbol{\Gamma} = (\gamma_{ij}),$ with $\gamma_{ij} = \Pr(g_t = j| g_{t-1} = i), \ i,j = 1, \ldots, N$. The connection of both stochastic processes arises from the assumption that the distribution of the observations $\boldsymbol{y}_t$ are fully determined by the currently active state, i.e.\ 
\begin{equation*}
f(\boldsymbol{y}_t|g_1, \ldots, g_T, \boldsymbol{y}_1, \ldots, \boldsymbol{y}_{t-1},\boldsymbol{y}_{t+1},\ldots,\boldsymbol{y}_T) = f(\boldsymbol{y}_t|g_t).
\end{equation*}
In general, $f$ can be any density or probability mass function depending on the type of data. Following the approaches of Franks et al. (2015), we opt for a Gaussian distribution.

>>>>

#### 2. Hidden Markov model

We describe the movements of the five defensive players during the phase of pre-snap motion by a hidden Markov model (see the Appendix for an in-depth description). In particular, we assume that each defender's $y$-coordinate at each time point $t$ is a realization from a particular Gaussian distribution with mean according to the $y$-coordinate of the offensive player that is currently guarded and an estimated standard deviation. HMMs are particularly useful in this setting as there is no direct knowledge of the guarding allocation. This is only partially observed based on the reaction of the defender to the offenders' movement. Naturally, the model thus treats the information on who is currently guarded as a latent state variable, and hence each observation as a realization from a mixture of Gaussian distributions (see Franks et al. 2015 for a similar approach in basketball).
Ultimately, the goal is to use the fitted model for so-called *state-decoding*, i.e. inferring information on the guarding allocation based on the observations made. For this task, HMMs not only leverage information on the $y$-coordinates of the defender and all offenders potetially guarded at each time point, but also incorporate probabilistic information based on the previous and subsequent observations. For practical details on the model fitting and state decoding, also see the Appendix.

@@

*Koslik J (2024). LaMa: Fast Numerical Maximum Likelihood Estimation for Latent Markov Models. R package version 2.0.2, <https://CRAN.R-project.org/package=LaMa>.

@@

#### Hidden Markov model

A hidden Markov model consists of an observed time series $\{\boldsymbol{y}_t\}_{t=1}^T$ and an unobserved first-order Markov chain $\{ g_t\}_{t=1}^T$, with $g_t \in \{1,\ldots,N\}$. In this case, at every time point $t$, $\boldsymbol{y}_t$ is the y-coordinate of the defensive player and $g_t$ proxies the offensives player to be guarded, i.e. the guarding allocation. 
The Markov chain is fully described by an initial distribution $\boldsymbol{\delta}^{(1)} = \bigl( \Pr(g_1=1), \ldots, \Pr(g_1=N) \bigr)$ and a transition probability matrix (t.p.m.) $\boldsymbol{\Gamma} = (\gamma_{ij})$, with $\gamma_{ij} = \Pr(g_t = j \mid g_{t-1} = i), \ i,j = 1, \ldots, N$. The connection of both stochastic processes arises from the assumption that the distribution of the observations $\boldsymbol{y}_t$ are fully determined by the state that is currently active. More formally,
\begin{equation*}
f(\boldsymbol{y}_t \mid g_1, \ldots, g_T, \boldsymbol{y}_1, \ldots, \boldsymbol{y}_{t-1},\boldsymbol{y}_{t+1},\ldots,\boldsymbol{y}_T) = f(\boldsymbol{y}_t \mid g_t = j), \qquad j \in \{1, \ldots, N\},
\end{equation*}
which we denote by $f_j(\boldsymbol{y}_t)$ in short.
In general, $f_j$ can be any density or probability mass function depending on the type of data and a typical choice is a parametric distribution with separate parameters for each latent state. Following the approaches of Franks et al. (2015), we opt for a Gaussian distribution with a mean that is fully determined by the current $y$-coordinate of defender $j \in \{1, \dots, N\}$ and a standard deviation that is fixed accross all states but estimated from the data.

To fit the model, we use direct numerical likelihood maximization. The HMM likelihood for the motion of a specific defender in a specific play can be calculated based on the so-called *forward algorithm*. It effectively performs a summation over all possible latent state sequences in an efficient manner, rendering the computational complexity linear in the number of observations. Time series of different defenders within the same play, and of different plays are treated as independent, hence their likelihood contributions are summed to obtain the full likelihood of the training data. For practical implementation, we wrote a custom likelihood function in R, using the function `forward()` and other convenience functions from the R package `LaMa` (Koslik, 2024) to speed up computations. Furthermore, we used the R package `RTMB` to make this likelihood function compatible with automatic differentiation, making the numerical optimization process more efficient and robust. 
The parameters to be estimated are only the transition probability matrix $\boldsymbol{\Gamma}$ and the standard deviation of the Gaussian distribution. 
The initial distribution of each defender for each play is fixed and computed based on ... ROUVEN

Having fitted an HMM to the data, we can use the model to predict the underlying state sequence based on the observations. This process is called state decoding and two main approaches exist. So-called *local decoding* constructs the conditional distributions
$$
\Pr(g_t = j \mid \boldsymbol{y}_1, \dots, \boldsymbol{y}_T)
$$
while *global decoding* using the *Viterbi-algorithm* finds the state sequence that maximizes the joint probability of the state sequence given the observations. Local decoding retains more probabilistic information as it provides a categorical state distribution for each time point while global decoding is more suitable for visualization purposes as it provides a single state sequence that is most likely to have generated the observations.
To obtain both the local state probabilities and the global state sequence, we used the functions `stateprobs()` and `viterbi()` that are also contained in the R package `LaMa`.
